---
title: "Paired Amplicon Analysis"
author: "Kat Terwelp"
format: pdf
editor: visual
---

## Paired Amplicon Analysis

Amplicon single nucleotide variants are paired using BLAST (looking for identical matches).

```{r}
# load libraries
library(phyloseq) # for working with phyloseq objects
library(qiime2R) # import qza files
library(microViz) # useful phyloseq functions
library(ggpubr) # add statistics to graphs
library(skimr) # create summaries of dataframes (super useful!)
library(vegan) # beta diversity analyses
```

```{r}
# import data 
cols <- c("query_asv", "query_len", "seq_asv", "seq_len", "alignment_len", "perc_id", "num_id", "evalue", "bitscore", "mismatch", "query_coverage", "gaps")

# import blast tsvs
blast_v1v3 <- read_tsv("../intermediates/paired-blast-results/FL16S_paired-blast-results.tsv", col_names = cols)
blast_fl <- read_tsv("../intermediates/paired-blast-results/V1V3_paired-blast-results.tsv", col_names = cols)

# import data and common utility functions
source("plot_utils.r", local = knitr::knit_global())

# import original dada2 sequences
v1v3path <- "../data/16S_V1-V3/dada2_seq.fasta.qza"
flpath <- "../data/FL-16S/dada2_seq.fasta.qza"
seq_v1v3 <- read_qza(v1v3path)
seq_fl <- read_qza(flpath)

# extract input ASVs for paired blast 
name_seq_fl <- names(seq_fl$data)
## 4036 for FL-16S 
name_seq_v1v3 <- names(seq_v1v3$data)
## 2497 for V1-V3

# functions
# function to filter blast table to just exact hits
f_filt_blast <- function(tbl) {
  tbl <- tbl %>% 
  filter(perc_id == 100) %>% 
  filter(alignment_len == seq_len | alignment_len == query_len) %>% 
  group_by(query_asv) %>% 
  mutate(num_query_matches = length(unique(seq_asv))) %>% # nASVs match query
  mutate(hit_type = "match") %>% 
    ungroup()
  return(tbl)
}


```

```{r}
# acquire match results
match_fl <- f_filt_blast(blast_fl)
match_v1v3 <- f_filt_blast(blast_v1v3)
match <- bind_rows(match_fl, match_v1v3)
```

```{r}
# add metadata information
# keep cols
meta_cols <- c("Sample", "Abundance", "shrt_sample_dna", "analysis", "total_reads", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "ASV")
# melt phyloseq object
melt <- all_ps %>% 
  ps_filter(analysis != "MGX") %>% 
  ps_melt() %>% 
  select(any_of(meta_cols)) %>% 
  rename_with(tolower) %>% 
  filter(abundance > 0) #remove nonexistant obs
melt_query <- melt %>% 
  rename_with(~paste0("query_", .x))
melt_seq <- melt %>% 
  rename_with(~paste0("seq_", .x))

# combine the metadata 
matched_meta <- full_join(match, melt_query) 
# full join to capture the "query" asvs that are not matches 
# or were filtered out in ps (aka in match but not melted phyloseq)
matched_meta <- left_join(matched_meta, melt_seq)
# left join to only get meta for sequences that matched queries
```

```{r}
# add information related to non-matches 
# (no identical hits from other platform)
matched_meta <- matched_meta %>% 
  mutate(hit_type = case_when(hit_type == "match" ~ "match", 
                              .default = "non-match")) 
# some ASVs were filtered out via taxonomy/decontam, these aren't in ps
# going to save these ASV names and add a metadata col to mark these
melt_asvs <- melt$asv %>% unique() # all asvs that were in the ps object
matched_meta <- matched_meta %>% 
  mutate(seq_filt_out = case_when(seq_asv %in% melt_asvs ~ FALSE, 
                                  .default = TRUE), 
         query_filt_out = case_when(query_asv %in% melt_asvs ~ FALSE, 
                                    .default = TRUE))
```

### Characterizing non-matches

**How many ASVs had matches in the other dataset?**

```{r}
# how many ASVs were matches for V1-V3 and FL? 
matched_meta %>% 
  filter(query_filt_out == FALSE) %>% # remove filtered ASVs
  select(query_asv, query_analysis, hit_type) %>% 
  distinct() %>% # remove duplicates
  group_by(query_analysis, hit_type) %>% 
  summarize(n = n()) %>% 
  group_by(query_analysis) %>% 
  mutate(total = sum(n), 
         percent = n/total*100) %>% 
  select(-total)
```

A greater proportion (\~66%) of V1-V3 ASVs have no FL-16S counterpart. In comparison, only \~54% of FL-16S ASVs have no V1-V3 counterpart.

**How abundant are matches vs non-matches?**

```{r}
matched_meta %>% 
  filter(query_filt_out == FALSE) %>% # remove filtered ASVs
  select(query_asv, query_abundance, query_shrt_sample_dna, hit_type, query_analysis) %>% 
  distinct() %>% 
  ggplot(aes(x = query_shrt_sample_dna, y = query_abundance, fill = hit_type)) +
      geom_bar(stat = "identity", position = "stack") +
      labs(title = "Relative Abundance of matched ASVs",
           x = "Sample",
           y = "Relative Abundance") +
      theme_minimal() +
  facet_wrap(~query_analysis) +
  coord_flip()

```

**For non-matches, what species are there?**

```{r}
matched_meta %>% 
  filter(query_filt_out == FALSE) %>% # remove filtered ASVs
  filter(hit_type == "non-match") %>% 
  select(query_asv, query_abundance, query_shrt_sample_dna, query_analysis, query_phylum) %>% 
  distinct() %>% 
  ggplot(aes(x = query_shrt_sample_dna, y = query_abundance, fill = query_phylum)) +
      geom_bar(stat = "identity", position = "stack") +
      labs(title = "Relative Abundance of unmatched ASVs",
           x = "Sample",
           y = "Relative Abundance") +
      theme_minimal() +
  facet_wrap(~query_analysis) +
  coord_flip()
```

```{r}
query_taxonomy <- matched_meta %>% 
  filter(query_filt_out == FALSE) %>% # remove filtered ASVs
  select(query_asv, query_analysis, hit_type, query_kingdom, query_phylum, query_class, query_order, query_family, query_genus, query_species) %>% 
  distinct()  
  

# who in FL-16S was not matched in V1-V3? 
unmatched_fl <- query_taxonomy %>% 
  filter(hit_type == "non-match" & query_analysis == "FL-16S") %>% 
  group_by(query_analysis, query_genus, query_species) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n))
head(unmatched_fl, n = 10)

# who in V1-V3 was not matched in FL-16S? 
unmatched_v1v3 <- query_taxonomy %>% 
  filter(hit_type == "non-match" & query_analysis == "V1-V3") %>% 
  group_by(query_analysis, query_genus, query_species) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n)) 
head(unmatched_v1v3, n = 10)
 
```

```{r}
query_
```
